---
title: "Final Report"
author: "Spencer Robinson, Thomas Sefton, and Paolo Spadano"
date: "2025-12-10"
output: 
  bookdown::pdf_document2:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggrepel)
library(xgboost)
library(recipes)
library(rsample)
library(tidymodels)
library(tictoc)
library(doParallel)
library(yardstick)
library(ranger)
```

# Introduction
University tuition in the United States shows a sharp contrast between public and private universities. Public institutions typically charge much lower in-state tuition because individual state governments subsidize a portion of post-secondary tuition for residents, while non-resident students at the same campuses have substantially higher sticker prices that more closely reflect the full cost of education. Private for-profit universities generally do not differentiate tuition costs based on students’ state residency but rather a set rate that is much higher on average than both in state and out-of-state prices at public institutions. National figures from the National Center for Education Statistics and College Board show that average tuition at private four-year colleges is roughly three to four times higher than in state tuition at public four-year universities, demonstrating the financial stakes of both institution type. 
Source: https://www.bestcolleges.com/research/average-cost-of-college/ https://nces.ed.gov/programs/coe/indicator/cua

Graduation outcomes also differ between public and private universities. National NCES data on six-year following graduation rates indicated that students at private four-year institutions graduate at slightly higher rates on average than those at public four-year colleges, a gap that other analyses attribute partly to differences in selectivity, resources, and student support services. For example, federal statistics report six-year graduation rates in the high 65-69 percent range at private compared with to a low 60-65 percent range at public institutions, while Urban Institute work using state-level samples finds that some private colleges within a state exceed 90 percent completion versus many public campuses clustered closer to 50 percent. Together, these patterns suggest that the same structural features that drive higher tuition at private universities may also be linked to stronger completion outcomes, even as considerable variation exists within each sector.
Sources:
https://nces.ed.gov/fastfacts/display.asp?id=40#:~:text=63%20percent%20at%20public%20institutions%2C%2068%20percent%20at%20private%20nonprofit 
https://www.bestcolleges.com/research/college-graduation-rates/

The structure of state higher-education systems further shapes how these patterns appear across the country. Nationwide tallies of Title IV institutions show that public universities make up less than one-third of all colleges, while private nonprofit and for-profit institutions together account for the majority, meaning that in many states most campuses are privately controlled even when public systems enroll more students. States with especially large higher-education sectors such as; New York, California, Texas, and Pennsylvania, tends to possess both a high number of institutions and a substantial share that are private, when looking at New York in particular, combining a dense network of private colleges with a sizable public system. As a result, plots relating the number of universities in each state to the proportion that are private often show an upward trend, with states that have more campuses overall also displaying a higher degree of privatization, though this relationship can vary and should be interpreted as descriptive rather than definitive evidence of a national rule.
Sources:https://nces.ed.gov/fastfacts/display.asp?id=1122
https://www.worldatlas.com/articles/which-state-has-the-most-colleges.html https://www.appily.com/colleges/state

Faculty compensation patterns mirror the classification of each tier of education achieved, with pay varying not only by rank but also by institution type. National faculty salary surveys show that professors at private, research-intensive universities typically earn more at every rank than their peers at less research-focused or two-year institutions, reflecting differences in revenue streams, endowment size, and competition for talent. 

# Research Question
What differentiates graduation rates for public and private colleges in the US? Additionally, can the type of college be judged without direct metrics like government funding or even tuition price? In answering these questions we also aim to explore the quantifiable differences in education quality between public and private institutions.

# Tidying
The tidying process involved reading two datasets: one on faculty compensation (AAUP) and one on college demographics and admissions (US News). Missing values (denoted by *) were converted to "NA" in both datasets. In the faculty compensation data, several columns representing averages and counts were multiplied by 100 to represent their real value. The US News data was improved by converting the numeric public/private code into descriptive categorical labels ("public" and "private"). Additionally, two new features were added by calculating the IQR for both SAT Math and SAT Verbal scores, which measures the spread of scores for admitted students. The data was then reformatted in two ways: first to create a summary table by grouping institutions by their public/private status and calculating mean tuition, graduation rate, and SAT scores. Second, to reshape the data from a wide format to a long format by consolidating all standardized test score variables (averages, Q1s, Q3s) into three descriptive columns (stat, test, and section) alongside a single score column.

```{r tidying}
aaup_url = "https://lib.stat.cmu.edu/datasets/colleges/aaup.data"
usnews_url = "https://lib.stat.cmu.edu/datasets/colleges/usnews.data"

aaup_cols = c("fed_id", "college_name", "state", "type", "avg_sal_full", "avg_sal_associate", "avg_sal_assistant", "avg_sal_all", "avg_comp_full", "avg_comp_associate", "avg_comp_assistant", "avg_comp_all", "num_full", "num_associate", "num_assistant", "num_instructor", "num_all")

usnews_cols = c("fed_id","college_name","state","public_private","avg_sat_math","avg_sat_verbal","avg_sat_combined","avg_act","q1_sat_math","q3_sat_math","q1_sat_verbal","q3_sat_verbal","q1_act","q3_act","num_apps_received","num_apps_accepted","num_new_enrolled","pct_top10_hs","pct_top25_hs","num_ft_undergrads","num_pt_undergrads","tuition_instate","tuition_outstate","room_board_cost","room_cost","board_cost","fees_additional","books_estimated","personal_spending_est","pct_faculty_phd","pct_faculty_terminal","student_faculty_ratio","pct_alumni_donate","instr_exp_per_student","grad_rate")

calc_iqr = function(q1, q3) q3-q1

aaup = read.csv(aaup_url, header = FALSE, col.names = aaup_cols, na.strings = "*") |>
  mutate(across(starts_with(c("avg", "num")), ~ .x * 100))

usnews = read.csv(usnews_url, header = FALSE, col.names = usnews_cols, na.strings = "*") |>
  mutate(public_private = factor(public_private, levels = c(1, 2), labels = c("public", "private"))) |>
  mutate(sat_math_iqr = calc_iqr(q1_sat_math, q3_sat_math), sat_verbal_iqr = calc_iqr(q1_sat_verbal, q3_sat_verbal))

train_val_split <- initial_split(usnews, prop = 0.8, strata = public_private)
uni_train_val <- training(train_val_split)
uni_test <- testing(train_val_split) # 20% for final evaluation

train_val_split_2 <- initial_split(uni_train_val, prop = 0.75, strata = public_private)
uni_train <- training(train_val_split_2) # 60% for training
uni_val <- testing(train_val_split_2) # 20% for validation

usnews <- uni_train # Ensures only training data is used for plots

pub_vs_priv = usnews |>
  group_by(public_private) |>
  summarize(avg_tuition = mean(tuition_instate, na.rm = TRUE), avg_grad_rate = mean(grad_rate, na.rm = TRUE), avg_sat = mean(avg_sat_combined, na.rm = TRUE))

sat_long <- usnews |>
  pivot_longer(
    cols = matches("^(avg|q1|q3)_(sat|act)"),
    names_to = c("stat", "test", "section"),
    names_pattern = "^(avg|q1|q3)_(sat|act)_?(math|verbal)?$",
    values_to = "score"
  ) |>
  select(fed_id, college_name, state, stat, test, section, score)

```


# Exploratory Plots
```{r}

# --- Plot 2: IQR Distribution Comparison (using usnews) ---
# Uses: public_private, sat_math_iqr, sat_verbal_iqr

iqr_long <- usnews |>
  select(public_private, sat_math_iqr, sat_verbal_iqr) |>
  pivot_longer(
    cols = ends_with("iqr"),
    names_to = "SAT_Section_IQR",
    values_to = "IQR_Value"
  ) |>
  # Clean up names for plot labels
  mutate(SAT_Section_IQR = str_remove(SAT_Section_IQR, "sat_"),
         SAT_Section_IQR = str_replace(SAT_Section_IQR, "_iqr", ""))

ggplot(iqr_long, aes(x = public_private, y = IQR_Value, fill = SAT_Section_IQR)) +
  geom_boxplot() +
  facet_wrap(~ SAT_Section_IQR) + # Separate Math and Verbal IQR distributions
  labs(
    title = "Interquartile Range (IQR) of SAT Scores by Institution Type",
    x = "Institution Type",
    y = "Score IQR (Q3 - Q1)",
    fill = "SAT Section"
  )
```


```{r plots}
usnews |>
  select(public_private, tuition_instate, tuition_outstate) |>
  pivot_longer(cols = c(tuition_instate, tuition_outstate), names_to = "instate_outstate", values_to = "tuition") |>
  ggplot() +
    geom_violin(aes(x = public_private, y = tuition, fill = instate_outstate, colour = instate_outstate)) +
  labs(
    title = "Distribution of In-State vs Out-of-State Tuition",
    subtitle = "Tuition Levels Differ Substantially Between Public and Private Institutions",
    x = "Institution Type",
    y = "Tuition Cost (USD)",
    fill = "Tuition Type",
    caption = "Private instututions charge far higher tuition, and out-of-state tuition exceeds in-state across both types.") + 
  theme_bw()
```
The distribution of tuition costs shows a clear divide between public and private institutions. Public universities offer significantly lower in-state tuition due to state subsidies, while out of state students face generally higher charges. Private universities do not differentiate based on residency status, display substantially higher tuition overall with greater variation. This pattern aligns with national tuition statistic from the U.S. Department of Education, which shows that private institutions charge substantially higher tuition and that out-of-state students at public universities pay considerably higher than in-state students (NCES, 2024). These differneces in tuition structue highlight the financial contrast between institution types and the importance of residency in determining affordability.
Source: https://nces.ed.gov/programs/digest/d23/tables/dt23_330.10.asp

```{r}
aaup_salary_long <- aaup |>
  mutate(
    type = factor(type)
    ) |>
  select(type, starts_with("avg_sal_")) |>
  pivot_longer(
    cols = starts_with("avg_sal_"),
    names_to = "rank",
    names_pattern = "avg_sal_(.*)",
    values_to = "avg_salary"
  ) |>
# I am ordering ranks for x-axis
  mutate(
    rank = factor(
      rank,
      levels = c("assistant", "associate", "full", "all"), 
      labels = c("Assistant", "Associate", "Full", "All ranks")
  ))
# Now summarizing mean salary by type and rank
salary_summary <- aaup_salary_long |>
  group_by(type, rank) |>
  summarise(
    mean_salary = mean(avg_salary, na.rm = TRUE) / 100,
    .groups = "drop"
  )

ggplot(salary_summary) +
  geom_col(aes(x = rank, y = mean_salary, fill = rank)) +
  facet_wrap(~type) +
  labs(
    title = "Average Faculty Salary by Rank and Institution Type",
    subtitle = "Faculty at Private Universities Tend to Earn More at Every Rank",
    x = "Faculty Rank",
    y = "Average Salary (USD)",
    caption = "Across AAUP institution types, average faculty salaries consistently rise with rank.") + 
  theme_bw()
```
These four bar plots compare average faculty salaries across AAUP institutional categories (I, IA, IIB, VIIB), showing that salaries consistently decline as institutions move from doctoral -granting, to two-year colleges. Faculty at Type I institutions earn the highest salaries across all ranks, reflecting the greater research intensity, funding capacity, and competitive hiring environment typical of doctoral universities. This aligns with AAUP's institutional classification system, which defines Type I institutions as research-intensive universities and Type IIB/VIIB institutions as primarily undergraduate or two year colleges with fewer graduate programs and lower average compensation structures (AAUP, 2023).
Source: https://www.aaup.org/explanation-statistical-data-5

```{r}
usnews_scatter <- usnews |> 
  group_by(state) |>
  summarize(universities = n(),
            percent_private = sum(public_private == "private") / n())

max_uni_state <- usnews_scatter |>
  filter(universities == max(universities))

ggplot(usnews_scatter) +
  geom_point(aes(x = universities, y = percent_private)) +
  geom_label_repel(data = max_uni_state, aes(x = universities, y = percent_private, label = state)) + 
  labs(
    title = "Relationship Beteen Number of Universities and % Private Institutions by State",
    subtitle = "Each Point Represents a U.S. State; the State with the Highest Number of Universities is Highlighted",
    x = "Number of Universities",
    y = "Proportion of Private Instututions",
    caption = "States with more universities tend to have higher proportion of private instututions, with New York as the outlier."
  ) + 
  theme_bw()
```
The scatter plot shows the relationship between the number of universities in each state and the proportion of those institutions that are private. The pattern suggests that states with larger higher-education systems tend to have a greater share of private institutions, with New York clearly standing as an extreme outlier due to both its high number of universities and its high concentration of private colleges. While this plot highlights a noticeable upward trend between system size and privatization within this data set, the observation is exploratory and reflects the structure of the data used in this analysis rather than a confirmed national pattern.

# Modelling Plots
```{r}
usnews |>
  mutate(state = fct_lump(state, 4)) |>
  ggplot() +
  geom_boxplot(aes(x = reorder(state, grad_rate, FUN=median), y = grad_rate, fill = state)) +
  facet_wrap(~public_private) +
  labs(
    title = "Graduation Rates Across States and Institution Types",
    subtitle = "States Grouped into the Top 4 Most Frequent Categories - Remaining States Grouped as 'Other'",
    x = "State (Top 4 + Other)",
    y = "Graduation Rate (%)",
    fill = "State",
    caption = "Private universities generally show higher median graduation rates than public across states.") +
  theme_bw()
```
This box and whisker plot compares graduation rates across the most common states in the dataset, seperated by public and private institutions. Private institutions consistently show higher median graduation rates than public universities within the same states, reflecting differences in selectivity student support and available resources. Findings from Urban Institute data, support our plot, showing that private four-year colleges report substantially higher graduation rates than public four-year colleges withing the same state, with private institutions reaching graduation rates above 90% compared to many public schools clustered near 50% (Urban Institute, 2024).
Source: https://www.urban.org/sites/default/files/publication/101635/comparing_colleges_graduation_rates_0.pdf

```{r}
usnews |>
  select(public_private, grad_rate, tuition_instate, tuition_outstate) |>
  pivot_longer(
    cols = c(tuition_instate, tuition_outstate),
    names_to = "tuition_type",
    values_to = "tuition"
  ) |>
  ggplot(aes(x = tuition, y = grad_rate, color = public_private)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  facet_wrap(~ tuition_type, scales = "free_x") + # Use free_x scale because tuition levels vary wildly
  labs(
    title = "Graduation Rate vs. Tuition Cost by Institution Type",
    subtitle = "Comparing In-State and Out-of-State Tuition Levels",
    x = "Tuition Cost (USD)",
    y = "Graduation Rate (%)",
    color = "Institution Type",
    caption = "Private institutions generally charge higher tuition but also show a wider range of graduation outcomes."
  ) +
  theme_bw()
```



```{r modelling}

# --- Linear Model Setup and Comparison ---

# Define the models and their alternative specifications
# Model 1: Graduation Rate vs. Instructional Expense per Student
grad_exp_lm_full <- linear_reg() |> set_engine("lm")
grad_exp_formula <- grad_rate ~ instr_exp_per_student * public_private # Continuous * Categorical (includes main effects)

# Model 2: Graduation Rate vs. Student-Faculty Ratio
grad_ratio_lm_full <- linear_reg() |> set_engine("lm")
grad_ratio_formula <- grad_rate ~ student_faculty_ratio * public_private # Continuous * Categorical (includes main effects)

# 1. Full Model (Model 1)
grad_exp_fit_full <- workflow() |>
  add_model(grad_exp_lm_full) |>
  add_formula(grad_exp_formula) |>
  fit(data = uni_train)

# 2. Full Model (Model 2)
grad_ratio_fit_full <- workflow() |>
  add_model(grad_ratio_lm_full) |>
  add_formula(grad_ratio_formula) |>
  fit(data = uni_train)

# --- Comparison Models for Model 1 ---

# A. Remove Continuous Feature (instr_exp_per_student)
grad_exp_no_cont <- workflow() |>
  add_model(grad_exp_lm_full) |>
  add_formula(grad_rate ~ public_private) |>
  fit(data = uni_train)

# B. Remove Interaction Term (instr_exp_per_student:public_private)
grad_exp_no_int <- workflow() |>
  add_model(grad_exp_lm_full) |>
  add_formula(grad_rate ~ instr_exp_per_student + public_private) |>
  fit(data = uni_train)

# --- Comparison Models for Model 2 ---

# C. Remove Continuous Feature (student_faculty_ratio)
grad_ratio_no_cont <- workflow() |>
  add_model(grad_ratio_lm_full) |>
  add_formula(grad_rate ~ public_private) |>
  fit(data = uni_train)

# D. Remove Interaction Term (student_faculty_ratio:public_private)
grad_ratio_no_int <- workflow() |>
  add_model(grad_ratio_lm_full) |>
  add_formula(grad_rate ~ student_faculty_ratio + public_private) |>
  fit(data = uni_train)

# --- Function to calculate RMSE on the test set ---

calculate_rmse <- function(fit, data) {
  preds <- predict(fit, new_data = data) |> 
    bind_cols(data |> select(grad_rate))
  
  rmse_value <- preds |> 
    rmse(truth = grad_rate, estimate = .pred) |>
    select(.estimate) |>
    pull()
  
  return(rmse_value)
}

rmse_full_exp <- calculate_rmse(grad_exp_fit_full, uni_val) 
rmse_no_cont_exp <- calculate_rmse(grad_exp_no_cont, uni_val) 
rmse_no_int_exp <- calculate_rmse(grad_exp_no_int, uni_val) 

rmse_full_ratio <- calculate_rmse(grad_ratio_fit_full, uni_val)
rmse_no_cont_ratio <- calculate_rmse(grad_ratio_no_cont, uni_val) 
rmse_no_int_ratio <- calculate_rmse(grad_ratio_no_int, uni_val) 

# --- Create Comparison Table ---

rmse_comparison_table <- tibble::tribble(
  ~Model, ~Model_Type, ~RMSE_Validation, # Changed header to reflect validation set usage
  "Grad_Exp_Full", "Full Model", rmse_full_exp,
  "Grad_Exp_No_Instr_Exp", "No Continuous Feature", rmse_no_cont_exp,
  "Grad_Exp_No_Interaction", "No Interaction Term", rmse_no_int_exp,
  "Grad_Ratio_Full", "Full Model", rmse_full_ratio,
  "Grad_Ratio_No_Stu_Fac_Ratio", "No Continuous Feature", rmse_no_cont_ratio,
  "Grad_Ratio_No_Interaction", "No Interaction Term", rmse_no_int_ratio
) |>
  mutate(Model_Family = if_else(str_detect(Model, "Exp"), "Model 1: Expense", "Model 2: Ratio"))

print(rmse_comparison_table)

ggplot(rmse_comparison_table) +
  geom_col(aes(x = Model, y = RMSE_Validation, fill = Model_Family), show.legend = FALSE) +
  geom_text(aes(x = Model, y = RMSE_Validation, label = round(RMSE_Validation, 2)), 
            vjust = 1) +
  facet_wrap(~ Model_Family, scales = "free_x") +
  labs(
    title = "Validation RMSE Comparison Across Linear Model Variations",
    subtitle = "RMSE calculated on the held-out validation set (uni_val)",
    x = "Model Specification",
    y = "Validation RMSE (Graduation Rate % Error)"
  ) +
  scale_x_discrete(labels = function(x) stringr::str_replace_all(x, "_", "\n")) + # Improve x-axis readability
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r advanced_modelling, results = 'hide'}


xgb_tune_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  mtry = tune(),
  min_n = tune(),
  learn_rate = 0.1,
) |>
  set_engine("xgboost") |>
  set_mode("classification")


xgb_params <- parameters(
  trees(range = c(500, 2000)),
  tree_depth(range = c(3, 10)),
  mtry(range = c(3, 30)),
  min_n(range = c(3, 40))
)

rf_tune_spec <- rand_forest(
  trees = 1000,
  mtry = tune(),
  min_n = tune()
) |>
  set_engine("ranger") |> 
  set_mode("classification")

rf_params <- parameters(
  mtry(range = c(3, 30)), 
  min_n(range = c(3, 40))
)

uni_folds <- vfold_cv(uni_train, v = 2, strata = public_private)

cl <- makePSOCKcluster(detectCores() - 4)
registerDoParallel(cl)
print("Parallel processing enabled")

full_recipe <- recipe(public_private ~ ., data = uni_train) |>
  update_role(
    c(fed_id, college_name), 
    new_role = "ID") |>
  step_novel(state) |>
  step_dummy(all_nominal_predictors(), one_hot = TRUE) |>
  step_impute_median(all_numeric_predictors()) |>
  step_normalize(all_numeric_predictors())

costless_recipe <- recipe(public_private ~ ., data = uni_train) |>
  update_role(
    c(fed_id, college_name, tuition_instate, tuition_outstate, room_board_cost, room_cost, board_cost, fees_additional, books_estimated, personal_spending_est, instr_exp_per_student), 
    new_role = "ID") |>
  step_novel(state) |>
  step_dummy(all_nominal_predictors(), one_hot = TRUE) |>
  step_impute_median(all_numeric_predictors()) |>
  step_normalize(all_numeric_predictors())

apps_recipe <- recipe(public_private ~ num_apps_received + num_new_enrolled + pct_top10_hs, data = uni_train) |>
  step_dummy(all_nominal_predictors(), one_hot = TRUE) |>
  step_impute_median(all_numeric_predictors()) |>
  step_normalize(all_numeric_predictors())

workflows <- workflow_set(preproc = list(full = full_recipe, costless = costless_recipe, apps = apps_recipe), 
                             models = list(xgb = xgb_tune_spec, rf = rf_tune_spec))


# Start timer
tic()

bayes_results = list()


map(workflows$wflow_id, function(id) {
  
  current_workflow <- workflows |>
        extract_workflow(id = id)
  
  current_metrics = metric_set(roc_auc, accuracy)
  
  if (str_detect(id, "rf")) {
    current_params <- rf_params
  } else {
    current_params <- xgb_params
  }
  
  
    # Run the Bayesian Optimization
  bayes_results[[id]] <<- tune_bayes(
    current_workflow,
    resamples = uni_folds,
    iter = 5,
    param_info = current_params,
    metrics = current_metrics,
    control = control_bayes(
      no_improve = 10,
      save_pred = TRUE,
      verbose = TRUE
  ))
})

# Stop parallel processing
stopCluster(cl)

# End timer
toc()

final_fits = list()
best_params = list()
for (mod_name in names(bayes_results)){

  current_best_params <- select_best(bayes_results[[mod_name]], metric = "roc_auc")
  
  
  best_params[[mod_name]] = current_best_params
  
  final_workflow <- workflows |>
    extract_workflow(id = mod_name) |>
    finalize_workflow(current_best_params)
  
  final_fits[[mod_name]] <- final_workflow |>
    fit(data = uni_train)
}


```

```{r}
final_results <- map_dfr(
  final_fits, 
  .f = function(model_fit) {
    
    test_predictions <- predict(model_fit, new_data = uni_test, type = "prob") |>
     bind_cols(uni_test |> select(public_private))

    class_predictions <- predict(model_fit, new_data = uni_test, type = "class")

    test_predictions <- test_predictions |>
      bind_cols(class_predictions)
      
    roc_auc_result <- test_predictions |>
      roc_auc(
        truth = public_private,
        .pred_public
      )
    
    accuracy_result <- test_predictions |>
      accuracy(
        truth = public_private,
        estimate = .pred_class 
      )
    
    bind_rows(roc_auc_result, accuracy_result)
  },
  .id = "model_name" 
)

final_results <- final_results |> 
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  separate_wider_delim(model_name, "_", names = c("recipe", "model"))

print(final_results)

ggplot(final_results) +
  geom_col(aes(x=model, y=accuracy, fill=model)) +
  facet_wrap(~recipe)
```
This more advanced model relates to the second goal of the study, attempting to predict whether a school is public or private.